Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-03-30 doc/invt.tex
intq.h: variate qt for the optimal schedule

We tested the above theory on a one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$.
%
%Since in the asymptotic regime,
%the distribution converges to $p_i$
%no matter the intrinsic distribution, $p^*_i$,
%we also set the latter to be flat for convenience.
For convenience,
the intrinsic distribution, $p^*_i$,
is a set to be flat as well.
%
Our goal is to observe how the fluctuating error
of the bias potential defined in Eq. \eqref{eq:error_sum}
depends on the schedule, $\alpha(t)$,
at the end of the simulation.

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and let the system undergo an equilibration period
of constant updating magnitude,
$\alpha(t) = \alpha_0$.
%
After this,
we reset the origin of the time, $t$,
and started the schedule specified by
either
Eq. \eqref{eq:q_opt}
or
Eq. \eqref{eq:alpha_invtlambda}.
%
We use the latter case to verify our theory.
%
We wish to see how the final error
depends on the parameter, $\lambda$,
and if it matches the predicted value.
%
Here, we used $t_0 = 1/(\lambda \, \alpha_0)$
to smooth the transition from equilibration.

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes,
a global one and a local one.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel, landau_binder}
to sample the state space:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
However,
in the first sampling scheme,
we choose the destination $j$
randomly out of the $n$ bins.
%
This sampling scheme,
referred to this as the global sampling scheme,
emulates the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}
in the long time limit.
%
In the second scheme,
we limit the destination $j$
to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme,
referred to this as the local sampling scheme,
models local sampling processes
typical in MD and straightforward MC simulations.
%

\subsection{Single-bin (WL) updating scheme}

We first tested the single-bin (WL) updating scheme
using the above protocol.
%
Here, we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:err_singlebin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling schemes,
the optimal $\lambda$ occurred around $1.0$,
the optimal value in the asymptotic regime.
%
Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
represents the relative updating magnitude.
%
Figure \ref{fig:err_singlebin} also shows that
a smaller than optimal value of $1/\lambda$
tends to increase the error more rapidly
than a larger than optimal value,
suggesting that the latter (over-updating)
is generally preferable to the latter (under-updating)
when the optimal value is unknown.

Another observation is that the two error functions
differed only by a multiplicative constant.
%
This again agreed with the theory.
%
Since all eigenvalues of the single-bin scheme are
the same, $\lambda_2 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling scheme only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

@@

We now turn to the nearest-neighbor updating scheme.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $t$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(t + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length, which is in agreement with
the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case, however,
the normalized error roughly maintained the sample shape,
and the optimal $\lambda$ is around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

To understand this dependence,
we observe that for shorter simulations,
long-wavelength modes
(with larger $\lambda_k$ and smaller $k$),
dominate the error function in Eq. \eqref{eq:error_invt}
because of the prefactor $\lambda_k \, \Gamma_k$,
which is inherited from
the equilibrium error under a fixed updating magnitude,
Eq. \eqref{eq:error_eql}.
%
For longer simulations,
short-wavelength modes
(with smaller $\lambda_k$ and larger $k$)
become more important,
for $\lambda_k$ also gives the decay rate
of the $k$th mode.
%and errors in smaller $\lambda_k$ modes
%are harder to damp out.
%
This drives the optimal $\lambda$ to shift toward
a smaller $\lambda_k$ value
as the simulation lengthens.
%
Note also that while increasing $\lambda$
also increases the error of the long-wavelength modes,
this rate of increase is much smaller,
as shown in the $\nu_k > 1$ branch
in Fig. \ref{fig:err_component}.
%
The shift of $\lambda$, however, is less noticeable
under the local sampling scheme,
because longer-wavelength modes more heavily
weighted by much larger values of $\Gamma_k$.

@@

\subsection{Gaussian (metadynamics) updating scheme}

The error of Gaussian updating scheme is similar to
that of the nearest neighbor scheme,
as shown in Fig \ref{fig:err_sig5}.
%
The dependency of the simulation length is more prominent.
%
Even for local sampling,
there is now a visible shift of the optimal $\lambda$.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsig5.pdf}
  }
  \caption{
    \label{fig:err_sig5}
    Normalized error, $(T + t_0) \, E$,
    versus the %proportionality constant,
    relative updating magnitude,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme with $\sigma = 5$ bins.
    %
    The results have been averaged over $1000$ independent runs.
  }
\end{center}
\end{figure}

\subsection{Accuracy of the inverse-time schedule}

In the above, we have assumed the inverse-time schedule,
Eq. \eqref{eq:alpha_invtlambda}.
%
In Fig. \ref{fig:alpha_cmp},
we compared the inverse-time schedule
with the optimal $\lambda$,
and the optimal schedule from integrating
Eq. \eqref{eq:q_opt}.
%
The difference was larger for the global sampling scheme.
%
As predicted by
Eqs. \eqref{eq:dinvadt_limit1} and \eqref{eq:dinvadt_limit2},
the rate of change of $1/\alpha(t)$
indeed increases from $\lambda_{\min}$
to $\bar \lambda \approx 1$.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/dinvacmp.pdf}
  }
  \caption{
    \label{fig:alpha_cmp}
    Optimal schedules, in terms of $\dot \alpha(t)$, from Eq. \eqref{eq:q_opt},
    versus the inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal parameter, $\lambda$.
    %
    For fair comparison,
    the $q(T) = \int_0^t \alpha(t) \, dt$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}

@@

\section{\label{sec:invt_alpha}
Inverse-time schedules}

The optimal schedule predicted
by the direct solution of Eq. \eqref{eq:q_opt}
is complicated for a general
multiple-bin updating scheme.
%
Several previous studies\cite{marsili2006, barducci2008, dickson2011}
have been focused on analytically simpler schedules.
%
In a similar spirit,
we consider here a variant of
the inverse-time formula, Eq. \eqref{eq:alpha_invt1}
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
with the proportionality constant, $\lambda$,
to be optimized.
%
We shall compute the error computed from
this variationally optimized schedule
and compare it with that from the optimal schedule.
%
We shall show that the optimal parameter, $\lambda$,
is sensitive to the simulation time, $T$,
which parallels the gradual shift of
the instantaneous eigenvalue,
$\lambda(t)$,
as discussed in Sec. \ref{sec:optschedule}.

>>>>

Our test system is
a model one-dimensional system
of $n = 100$ bins,
with a flat target distribution
$p_i = 1/n$
under the reflective boundary condition.
%
%Since in the asymptotic regime,
%the distribution converges to $p_i$
%no matter the intrinsic distribution, $p^*_i$,
%we also set the latter to be flat for convenience.
For convenience,
the intrinsic distribution, $p^*_i$,
is a set to be flat as well.
%
Our goal is to observe how the fluctuating error
of the bias potential defined in Eq. \eqref{eq:error_sum}
depends on the schedule, $\alpha(t)$,
at the end of the simulation.

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and equilibrate the system under
a constant updating magnitude,
$\alpha(t) = \alpha_0$.
%
After this,
we reset the origin of the time, $t$,
and started the target schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

For the Gaussian updating scheme,
the kernel is truncated at
$\min\{10 \, \sigma, n - 1\}$
and numerically stabilized
using the technique
in Sec. \ref{sec:stabilize_wband}.

Since the error depends on
the underlying sampling process,
we have devised two sampling processes,
a global one and a local one.
%
Both cases were based on
Metropolis algorithm\cite{metropolis1953, newman, frenkel,
landau_binder}:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
However,
in the first and global sampling process,
we choose the destination $j$
randomly out of the $n$ bins.
%
This sampling process
emulates the perfect sampling process
discussed in Sec. \ref{sec:Gamma}
in the long time limit.
%
In the second or local process,
we limit the destination $j$
to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling process
in Sec. \ref{sec:Gamma}.
%

\subsection{Inverse-time schedule}

We first tested the theory using
the inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a parameter to be optimized.
and $t_0 = 1/(\lambda \, \alpha_0)$
to smooth the transition from equilibration.
%
This schedule is analytically simpler
than the optimal schedule,
and it is studied in Appendix \ref{sec:invt_schedule}.
%
We wish to see how the final error
depends on the parameter, $\lambda$,
and if it matches the predicted value,
given by Eq. \eqref{eq:error_invt}.

\subsubsection{Single-bin (WL) updating scheme}

For the single-bin updating scheme,
we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:err_singlebin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling processes,
the optimal $\lambda$ occurred around $1.0$,
verifying Eq. \eqref{eq:alpha_invt}.
%
%Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
%represents the relative updating magnitude.
%%
%Figure \ref{fig:err_singlebin} also shows
%an asymmetry of the error curve.
%A smaller than optimal value of
%the updating magnitude, $1/\lambda$,
%tends to increase the error more rapidly
%than a larger than optimal value,
%suggesting that the over-updating
%is generally preferable to the under-updating
%when the optimal value is unknown.

We also observe
that the two error functions
differed only by a multiplicative constant,
again again agreeing with the theory.
%
With identical eigenvalues,
$\lambda_1 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling process only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

@@

\subsubsection{Nearest-neighbor updating scheme}

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \eqref{sec:nnscheme}.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $T$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(T + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length, $T$,
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length, which is in agreement with
the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case,
the $T$-dependence was hardly visible
because of the predominantly large $\Gamma_k$'s
of the small $k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

%To understand this dependence,
%we observe that for shorter simulations,
%long-wavelength modes
%(with larger $\lambda_k$ and smaller $k$),
%dominate the error function in Eq. \eqref{eq:error_invt}
%because of the prefactor $\lambda_k \, \Gamma_k$,
%which is inherited from
%the equilibrium error under a fixed updating magnitude,
%Eq. \eqref{eq:error_eql}.
%%
%For longer simulations,
%short-wavelength modes
%(with smaller $\lambda_k$ and larger $k$)
%become more important,
%for $\lambda_k$ also gives the decay rate
%of the $k$th mode.
%%and errors in smaller $\lambda_k$ modes
%%are harder to damp out.
%%
%This drives the optimal $\lambda$ to shift toward
%a smaller $\lambda_k$ value
%as the simulation lengthens.
%%
%Note also that while increasing $\lambda$
%also increases the error of the long-wavelength modes,
%this rate of increase is much smaller,
%as shown in the $\nu_k > 1$ branch
%in Fig. \ref{fig:err_component}.
%%
%The shift of $\lambda$, however, is less noticeable
%under the local sampling scheme,
%because longer-wavelength modes more heavily
%weighted by much larger values of $\Gamma_k$.

@@

%\subsection{Gaussian (metadynamics) updating scheme}
%
%
%The error of Gaussian updating scheme is similar to
%that of the nearest neighbor scheme,
%as shown in Fig \ref{fig:err_sig5}.
%%
%The dependency of the simulation length is more prominent.
%%
%Even for local sampling,
%there is now a visible shift of the optimal $\lambda$.
%
%
%\begin{figure}[h]
%\begin{center}
%  \makebox[\linewidth][c]{
%    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsig5.pdf}
%  }
%  \caption{
%    \label{fig:err_sig5}
%    Normalized error, $(T + t_0) \, E$,
%    versus the %proportionality constant,
%    relative updating magnitude,
%    $1/\lambda$,
%    in Eq. \eqref{eq:alpha_invtlambda}
%    for the Gaussian updating scheme with $\sigma = 5$ bins.
%    %
%    The results have been averaged over $1000$ independent runs.
%  }
%\end{center}
%\end{figure}
%

\subsection{Optimal schedule}

We now turn to the optimal schedule
predicted by Eq. \eqref{eq:q_opt}.
%
For the Gaussian updating scheme
with $\sigma = 10$,
we computed the schedule,
and compared it with the inverse-time schedule
Eq. \eqref{eq:alpha_invtlambda}
with the optimal $\lambda$.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt},
    versus the optimized inverse-time schedule
    Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    %
    $q(T) = \int_0^t \alpha(t) \, dt$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}

As shown in Fig. \ref{fig:optacmp}(a),
the optimal schedule has a more complex curvature
than the optimized inverse-time schedule.
%
In Fig. \ref{fig:optacmp}(b),
we show that in the former case,
the rate of change of $\alpha^{-1}(t)$
increases gradually
from $\lambda(0)$ to $\lambda(T) \approx 1$,
as discussed in Sec. \ref{sec:optschedule}.
%
The errors in the two cases were
$4.0 \times 10^{-7}$
and
$5.9 \times 10^{-7}$,
respectively,
for the global sampling process,
%
or
$1.67 \times 10^{-4}$
and
$1.26 \times 10^{-4}$,
respectively
for the local sampling process,
suggesting a slight edge of
the optimal schedule from Eq. \eqref{eq:q_opt}.

\subsection{Comparison of updating schemes}

@@

\section{\label{sec:invt_schedule}
Inverse-time schedules}

The optimal schedule predicted
by the direct solution of Eq. \eqref{eq:q_opt}
is complicated for a general
multiple-bin updating scheme.
%
Several previous studies\cite{marsili2006, barducci2008, dickson2011}
have been focused on analytically simpler schedules.
%
In a similar spirit,
we consider here
the inverse-time formula,
Eq. \eqref{eq:alpha_invtlambda},
%
with the proportionality constant, $\lambda$,
to be optimized.
%
We shall compute the error computed from
this variationally optimized schedule
and compare it with that from the optimal schedule.
%
We shall show that the optimal parameter, $\lambda$,
is sensitive to the simulation time, $T$,
which parallels the gradual shift of
the instantaneous eigenvalue,
$\lambda(t)$,
as discussed in Sec. \ref{sec:optschedule}.
